---
# Below block of code was implemented based on review suggestion from Isuue 1.0

title: "Explore And Summarize Data: Red Wine Quality"
subtitle: "for **AIRBUS** Data Analyst Nanodegree Program"
author: "Pankaj NATH"
date: "`r format(Sys.time(), '%d %B, %Y')` Issue: 2.0"

output: 
  html_document:
    # theme of html document
    # theme of code highlight                                 
    # table of contents
    theme       : cosmo            # "default", "cerulean", "journal",
                                    # "flatly", "readable", "spacelab",
                                    # "united", "cosmo", "lumen", "paper", 
                                    # "sandstone", "simplex", "yeti"
    highlight   : tango          # "default", "tango", "pygments",
                                    # "kate",  "monochrome", "espresso",
                                    # "zenburn", "haddock", "textmate"
    toc         : true              # get table of content
    toc_depth   : 3
    toc_float   : true
    code_folding: hide
    number_sections: yes
    fig_caption: yes
    df_print    : paged

---

______________

# Table of Contents
1. Dataset Introdution
2. Univariate Plots
3. Univariate Analysis
4. Bivariate Plots
5. Bivariate Analysis
6. Multivariate Plots
7. Multivariate Analysis
8. Final Plots and Sumary
9. Reflection


# Dataset Introduction
The dataset opted for this project to explore and summarize is **"Red Wine
Quality"** dataset of the Portuguese **"Vinho Verde"** wine. This can be downloaded 
by clicking [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv).
The dataset has 1599 observations. But due to privacy and logistic issues, 
only physicochemical (inputs) and sensory (the output) variables are available 
(e.g. there is no data about grape types, wine brand, wine selling price, etc.)

Based on physicochemical tests, the attributes are listed below:

1. **fixed acidity** (tartaric acid - g / dm^3) : most acids involved with wine or 
fixed or nonvolatile (do not evaporate readily).

2. **volatile acidity** (acetic acid - g / dm^3) : the amount of acetic acid in 
wine, which at too high of levels can lead to an unpleasant, vinegar taste.

3. **citric acid** (g / dm^3) : found in small quantities, citric acid can add 
'freshness' and flavor to wines.

4. **residual sugar** (g / dm^3) : the amount of sugar remaining after fermentation 
stops, it's rare to find wines with less than 1 gram/liter and wines with 
greater than 45 grams/liter are considered sweet.

5. **chlorides** (sodium chloride - g / dm^3) : the amount of salt in the wine.

6. **free sulfur dioxide** (mg / dm^3) : the free form of SO2 exists in equilibrium 
between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents 
microbial growth and the oxidation of wine.

7. **total sulfur dioxide** (mg / dm^3) : amount of free and bound forms of S02; in 
low concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine.

8. **density** (g / cm^3) : the density of water is close to that of water depending
on the percent alcohol and sugar content.

9. **pH** : describes how acidic or basic a wine is on a scale from 0 (very acidic) 
to 14 (very basic); most wines are between 3-4 on the pH scale.

10. **sulphates** (potassium sulphate - g / dm3) : a wine additive which can 
contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial 
and antioxidant.

11. **alcohol** (% by volume) : the percent alcohol content of the wine.

Based on sensory data, the output variable:

12. **quality** : score between 0 (very bad) and 10 (very excellent).

This would be interesting to see how each attribute contribute towards wine's 
quality and also to see how these attributes are related with each other. Also 
we will investigate using available data if their are strong indications that 
wine quality depends on any particular chemical properties from above list.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading all of the packages to be used during this analysis
library(ggplot2)
library(gridExtra)
library(reshape2)

# Below lines of code were added based on Suggestion from review of Issue 1.0
# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9, fig.height=5, fig.path='Figs/',
                      fig.align='center', tidy=TRUE,
                      echo=FALSE, warning=FALSE, message=FALSE)
```

Lets' summarize our red wine dataset and view the variables and their mean, 
median, min, max and 1st & 3rd quartile values.

```{r echo=FALSE, Load_the_Data}
# Load the Data
red_wine <- read.csv('wineQualityReds.csv')

# Get summary of valiables in dataset
summary(red_wine)

# Dropping the first id/numbering column
red_wine <- subset(red_wine, select = -c(X))
```

# Univariate Plots Section

## Wine Quality
Let's start by exploring wine quality which ranges in [`r range(red_wine$quality)`] for our dataset.
This is definately an ordered categorical variable.
```{r echo=FALSE, Univariate_Plot_quality}
# plotting histogram for wine quality
ggplot(red_wine, aes(quality)) +  geom_histogram(binwidth = 1)

# Summarizing the quality variables value
summary(red_wine$quality)

# tabling the quality variable by count in each category of rating value
table(red_wine$quality)
```

There are only 10 observations with quality rating of 3 and 18 observations 
with quality rating of 8. Majority of observations are rated at 5 or 6.
Distribution of quality is normal but on a very narrow range.

```{r, echo=FALSE, message=FALSE, warning=FALSE, plots}
# Plotting histogram for all other variables
p1 <- qplot(red_wine$fixed.acidity)
p2 <- qplot(red_wine$volatile.acidity)
p3 <- qplot(red_wine$citric.acid)
p4 <- qplot(red_wine$residual.sugar)
p5 <- qplot(red_wine$chlorides)
p6 <- qplot(red_wine$free.sulfur.dioxide)
p7 <- qplot(red_wine$total.sulfur.dioxide)
p8 <- qplot(red_wine$density)
p9 <- qplot(red_wine$pH)
p10 <- qplot(red_wine$sulphates)
p11 <- qplot(red_wine$alcohol)
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, ncol = 4)
```

It would be interesting to investigate distribution of Citric Acid, Resedual 
Sugar, Chlorides, Free Sulphur Dioxide, Total Sulphur Dioxide, Sulphate and 
Alcohol. All of these variables do not seem to be normally distributed.

```{r, echo=FALSE, myFunc}
# Formatted this function by self in Issue 2.0
# Defining a function for plotting which we will call few times

# my_plot() : Function
## Output: Histogram
## Input: x - a variable for x-axis
##        y - a variable for y-axis
my_plot <- function(x, y) {
  ggplot(red_wine, aes(x)) +  geom_histogram(binwidth = y)  +
  geom_vline(xintercept = median(x), color = 'blue') +
  geom_vline(xintercept = mean(x), color = 'red')
}
```


## Alcohol
```{r echo=FALSE, Univariate_Plots_alcohol}
# plotting histogram for alcohol quantity
my_plot(red_wine$alcohol, 0.5)

summary(red_wine$alcohol)
```
Clearly the alcohol quantity in our red wine dataset is skwed with median
of 10.2 being closer to minimum (8.4) compared to maximum (14.9).

## Sulphates
```{r echo=FALSE, Univariate_Plots_sulphates}
# plotting histogram for sulphates
my_plot(red_wine$sulphates, 0.1)
summary(red_wine$sulphates)
```
Distribution of sulphates is also skewed and mean & median is very close 
to minimum with higher gap between third quartile and maximum. It would be 
interesting to investigate the quality of these outliers on the higher side.

## Chlorides
```{r echo=FALSE, Univariate_Plots_chlorides}
# plotting histogram for Chlorides quantity
my_plot(red_wine$chlorides, 0.01)
summary(red_wine$chlorides)
```

Distribution of chlorides is skewed similar to sulphates, with outliers on the higher side.

## Residual Sugar
```{r echo=FALSE, Univariate_Plots_sugar}
# plotting histogram for Residual Sugar quantity
my_plot(red_wine$residual.sugar, 1)
summary(red_wine$residual.sugar)
```

Distribution of residual sugar is skewed similar to sulphates and chlorides 
with outliers on the higher end.

## Citric Acid
```{r echo=FALSE, Univariate_Plots_citric}
# plotting histogram for citric acid quantity
my_plot(red_wine$citric.acid, 0.001)
summary(red_wine$citric.acid)
```

With a smaller bin size of 0.001 we can observe very interesting fact that the 
citric acid quantity in our dataset is clustered. It would be interesting to 
see if quality is also clustered with it or not?

## Total Sulphur Dioxide
This is the sum of free sulphur dioxide and bound sulphur dioxide. Using 
total SO2 along with free SO2 may have correlation between them. To break 
any bias between them, a new variable bound  sulphur dioxide will be created 
and total SO2 will be dropped from the dataset.

```{r echo=FALSE, Univariate_Plots_so2}
# Creating new variable bound.sulfur.dioxide
red_wine$bound.sulfur.dioxide <- red_wine$total.sulfur.dioxide - red_wine$free.sulfur.dioxide

# Dropping variable total.sulfur.dioxide from dataset
red_wine <- subset(red_wine, select = -c(total.sulfur.dioxide))
```

## Free & Bound Sulphur Dioxide
```{r echo=FALSE, Univariate_Plots_so2breakup}
# plotting histogram for free.sulfur.dioxide quantity
my_plot(red_wine$free.sulfur.dioxide, 1)
summary(red_wine$free.sulfur.dioxide)
my_plot(red_wine$bound.sulfur.dioxide, 5)
summary(red_wine$bound.sulfur.dioxide)
```

Both Free and Bound Sulphur dioxide is also has a skewed long tailed
distribution with outliers like previous other variables.


# Univariate Analysis

## What is the structure of your dataset?
> The red wine dataset has 12 parameters including quality with a total of 
1599 observations or datapoints. Quality is an ordered categorical parameter 
whereas rest all were of type float. This dataset was a tidy one hence no 
values were missing and no cleaning was required except dropping the X 
parameter which was nothing but id for each observation point in dataset and 
not needed for my analysis.

## What is/are the main feature(s) of interest in your dataset?
> Main feature of this dataset is the quality which has defined range of 0 
(very poor) to 10 (very excellent) but our dataset has values in the range of 
[`r range(red_wine$quality)`] only. I would like to investigate how each of 
the other features affect quality.

## What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
> Most of the features are skewed in distribution but quality has most 
observations at rating 5 or 6. With just Univariate Plots, it is difficult to 
predict or say anything of sure. I would like to go ahead and continue with 
Bivariate and Multivariate analysis.

## Did you create any new variables from existing variables in the dataset?
> Total Sulfur Dioxide was sum of Free and Bound Sulfur Dioxide in the wine. 
Our dataset had both total and Free SO2 values present. Since these two 
features are directly related to each other and are biased, I decided to create
and new variable Bound Sulfur Dioxide using these two and then drop Total 
sulfur Dioxide from the dataset. This gave me two variables which were no more directly related.

## Of the features you investigated, were there any unusual distributions?
> One interesting distribution of Citric Acid was observed to be clustered. There were many local peaks present. 

## Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
> As mentioned in previous questions, with justifications, X and Total Sulfur 
Dioxide variables were dropped and a new variable Bound Sulfur Dioxide was 
created. 

# Bivariate Plots Section

Let us begin with investigating correlations between all the variables using a 
correlation matrix. Based on the value of Pearson's coefficient of correleation
we can say any value less than +/- 0.4 shows weak or no correlation. Any +/- 
value between 0.5 and 0.7 shows moderate correlation. And any value higher to 
+/- 0.7 shows strong correleation.

```{r echo=FALSE, Bivariate_Plots}
cormat <- round(cor(red_wine), 2)
cormat
```

We didnot observed any stong correlation between any of the variables but yes 
some moderate correlation is observed between following pairs of variables:

1. fixed.acidity & citric.acid : 0.67

2. fixed.acidity & density : 0.67

3. fixed.acidity & pH : -0.68

4. volatile.acidity & citric.acid : -0.55

5. citric.acid & pH : -0.54

6. density & alcohol : -0.50

Let's also visualize the heatmap for our correlation matrix below.
The code used for this visualization is taken from  [here](http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization).
```{r echo=FALSE, heatmap}
# This block of code is updated by self in Issue 2.0

# Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
  cormat[lower.tri(cormat)]<- NA
  return(cormat)
}

# Function to reorder correlation matrix
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}

# Reorder the correlation matrix
cormat <- reorder_cormat(cormat)

# Get the upper triangle of correlation matrix
upper_tri <- get_upper_tri(cormat)

# Melt the correlation matrix
melted_cormat <- melt(upper_tri, na.rm = TRUE)

# Create a ggheatmap
ggheatmap <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson Correlation\nMatrix") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()

ggheatmap + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 2.5) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.7),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))
```

## Plot of fixed.acidity & citric.acid
Both these variables are showing a positive correlation of **0.67**. Lets view 
them in a scatter plot with trendline.
```{r echo=FALSE, Bivariate_Plots_1}
ggplot(red_wine, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm")
```

## Plot of fixed.acidity & pH
Both these variables are showing a negative correlation of **-0.68**. Lets view
them in a scatter plot with trendline.
```{r echo=FALSE, Bivariate_Plots_2}
ggplot(red_wine, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm")
```

With Univariate Plots/Analysis we observed that distribution of Sulphates, 
Chlorides and Citric Acid was all skewed with long tailed towards the right. 
These three variables are also having very weak correlation between them. 
Lets see how these three variables vary with quality.

```{r, echo=FALSE, myFunc2}
# This function is also updated following suggestions from review of Issue 1.0

# Defining a function for plotting which we will call few times
# my_plot2() : Function
## Output: Scatterplot (jittered)
## Input: x1 - a discrete variable
##        y1 - a continuous variable
##         z - alpha level
my_plot2 <- function(x1, y1, z) {
  ggplot(red_wine, aes(x = x1, y = y1)) +
  geom_jitter(width=0.35, height = 0, alpha = z) +
  geom_smooth(method = "lm")
}
```

## Plot of quality vs sulphates
```{r, echo=FALSE, Bivariate_Plots_3}
my_plot2(red_wine$quality, red_wine$sulphates, 0.25)
```

As expected, we cannot comment on quality based on sulphates quantity in red wine.

## Plot of quality vs chlorides
```{r, echo=FALSE, Bivariate_Plots_4}
my_plot2(red_wine$quality, red_wine$chlorides, 0.25)
```

From the plot the weaker correlation between quality and chlorides is evident. 
Lower chlorides in red wine doesnot gives higher quality of red wine.

## Plot of quality vs alcohol
```{r, echo=FALSE, Bivariate_Plots_5}
my_plot2(red_wine$quality, red_wine$alcohol, 0.25)
```

The correlation coefficient between alcohol quantity and quality of red wine 
was **0.48**. It wasn't much but the trendline shows that the most high quality
red wines have higher alcohol quantity in them. A boxplot between quality and 
alcohol variables may provide some more insights.

```{r echo=FALSE, message=FALSE, Bivariate_Plots_6}
# Setting the continuous quality variable to ordered
red_wine$quality <- factor(red_wine$quality, ordered=TRUE)

# boxplot between quality & alcohol
ggplot(red_wine, aes(x=quality, y=alcohol)) +
  geom_boxplot()
```

# Bivariate Analysis

## Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
> The feature of interest for our examination is quality. But we did not find 
any strong correlation between any of the other features with our feature of 
interest (quality). Quality had near to moderate positive correlation of 
**0.48** with alcohol and highest negative correlation of **-0.39** with 
volatile acidity. We validated the weaker trend that with sulphates and 
chlorides.

## Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
> The very first action was to calculate the correlation coefficients 
between all features involved. Overall we did not find any strong correlations 
(> 0.7) between these features but some moderate correlations (> 0.5 & < 0.7) 
were found. Among them, two pairs of features had the highest positive 
correlation of **0.67** between fixed.acidity & citric.acid and fixed.acidity
& density. The highest negative correlation on **-0.68** was found for 
fixed.acidity & pH.


## What was the strongest relationship you found?
> The only strongest positive trend we saw for quality was with alcohol where 
most of the higher rated wines had higher alcohol content. With the help of 
boxplot we found that initially the alcohol content drops from a poor quality 
wine to moderate quality (rating 5). Afterwards the alcohol content increases 
with quality. But this correlation doesnot prove this causation since many of 
the higher alcohol content wines were rated moderately. 

# Multivariate Plots Section

Based on feature pairs with higher correlation coefficients, I selected citric
acid and alcohol content feature for further exploration since these have 
higher correlation with quality feature. This is classic feature selection
method.

First multivariate plot will include Citric Acid, Alcohol and Quality.

```{r echo=FALSE, Message=FALSE, Multivariate_Plots_1}
ggplot(red_wine, aes(x = citric.acid, y = alcohol, color = quality)) +
  geom_jitter()
```

The features selected for first multivariate plot do not convey any strong
message. But a slight hint is visible that moderate quality red wines have
alcohol content and citric acid values on the lower end, whereas quality 
improves with increase in these two featrue values.

Second multivariate plot will include Alcohol, Density and Quality.

```{r echo=FALSE, Message=FALSE, Multivariate_Plots_2}
ggplot(red_wine, aes(x = alcohol, y = density, color = quality)) +
  geom_jitter() + geom_point(alpha = 0.005) + geom_smooth(method = "lm")
```

Second multivariate plot shows that more better quality red wines have 
alcohol content above 10 and density lower to 0.9975.

# Multivariate Analysis

## Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
> Higher alcohol and citric acid content combined with lower density are essential but not sufficient features for better quality of red wine.

## Were there any interesting or surprising interactions between features?
> The relationship between alcohol and density: increase in alcohol quantity 
and decrease in density, is interesting for me to investigate if this can 
really be possible in terms of chemistry?

------

# Final Plots and Summary

## Plot One
```{r echo=FALSE, Plot_One}
# Plot One is changed based on feedback recieved on Issue 1.0

# Setting the ordered quality variable to continuous
#red_wine$quality <- factor(red_wine$quality, ordered=FALSE)
red_wine$quality <- as.numeric(as.character(red_wine$quality))

ggplot(red_wine, aes(quality)) +  
  geom_histogram(binwidth = 1, color="darkblue", fill="lightblue") + 
  xlab("Wine Quality Rating (0- Poor; 10- Excellent)") +
  ylab("Count") +
  ggtitle("Histogram of Wine Quality") + 
  geom_vline(xintercept = median(red_wine$quality), color = 'black') +
  geom_vline(xintercept = mean(red_wine$quality), color = 'red') +
  geom_text(aes(x=median(red_wine$quality), label="\nMedian", y=200), colour="black", angle=90, size=5) +
  geom_text(aes(x=mean(red_wine$quality), label="\nMean", y=200), colour="red", angle=90, size=5)

# Setting back the continuous quality variable to ordered
red_wine$quality <- factor(red_wine$quality, ordered=TRUE)
```

### Description One
The red wine dataset had 12 features including the feature of our interest (quality). The first plot is a histogram showing the distribution of quality in our red wine dataset. It was evident from this plot that most of our wine samples were from rating 5 or 6. This dataset did not have any outliers with rating 0 or 10 with a mean of **5.636** and median of **6**. This dataset had very narrow distribution range of quality as [**3, 8**].

## Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(red_wine, aes(x=quality, y=alcohol, fill=quality)) + geom_boxplot() + 
  xlab("Wine Quality Rating (0- Poor; 10- Excellent)") +
  ylab("Alcohol Concentration (% by volume)") +
  ggtitle("Relationship between Alcohol Concentration and Wine Quality")
```

### Description Two
We identified that Alcohol Content as a feature had a strong relationship with
Quality compared to all other features. This second plot which is a boxplot 
then help us understand how the mean of Alcohol concentration is varying in
different wine qualities. We understand from this boxplot that highly rated red
wines have higher alcohol concentration by volume.

## Plot Three
```{r echo=FALSE, Plot_Three}
# Color encoding in this block of code is updated based on review feedback 
# from Issue 1.0

ggplot(red_wine, aes(x = alcohol, y = density, color = quality)) +
  geom_point(alpha = 0.25) + geom_smooth(method = "lm") +
  xlab("Alcohol Concentration (% by volume)") +
  ylab("Density (g / cm^3)") +
  labs(color='Quality') + 
  ggtitle("Variation in Alcohol Concentration and Density for different Quality") + 
  scale_color_brewer(labels = c("3- Poor", "4", "5", "6", "7", "8- Excellent"),
                     palette = "RdYlBu", direction = -1, name = "Quality") +
                    theme(panel.grid.minor = element_blank(), 
                    panel.background = element_rect(fill = "lightgrey"))
```

### Description Three
The third plot integrates density feature along with alcohol concentration to
see the behaviour with respect to quality of wine. It is then visible that 
highly rated red wines have lower density and higher alcohol concentration by volume.

-------------------------------------------------------------------------------

# Reflection
In this project I explored and examined a tidy red wine dataset which was
composed of 1599 observations of 12 variables. These variables included
physicochemical properties of red wine and median of quality from three human
expert testers.

I started my investigations firstly by reading the information about the 
dataset and its variables. Then I started with analyzing the distributions of
each of these variables in univariate plots and analysis. Before starting
bivariate analysis, I did the heatmap of Pearson's correlation coefficients.
This was helpful in picking up pairs of variables to bivariate plots and
analysis. Then I concluded my multivariate plots and analysis.

It was very interesting to see the distribution of citric acid which had many
local maximums and minimums. I was wondering if citric acid is really an
important variable or not? Though it had only a weak correlation with quality
variable, I wonder having data domain knowledge would have been very helpful.

The first challenge I faced while initiating this investigation was to identfy
and install all the packages that I would need. My computer was always 
throughing error. Then I came across the [R Studio Cloud](https://rstudio.cloud),
it was very smooth in working and all my work was in cloud, accessible from
anywhere, anytime. I loved working on this platform.

While working with the dataset, I identified that quality was a categorical
variable. So I changed it to an ordered variable. In some plots I faced issues
since it became discrete variable. I also then thought about why can't this
quality variable be a continuous? I found the answer, it was a median from 
three wine testers who may have rated only using integers between 0 to 10.

A lot could have been possible to investigated if the outcome of each wine 
tester was provided separately. Then it could have been the analyst's choice 
to take a continous mean or median to proceed with.

Also this dataset had many observations from rating 5 or 6. It was missing 
some outliers such a very poor (rating 0) or very excellent (rating 10). 
Without these outliers we may have missed the actual relationship these best 
or worst red wine would have with all other 11 variables.

I did not created any model with this dataset. With above mentioned deficiency
of this dataset, I wonder if the model will work good or not. May be for 
future this dataset can be imroved then creating a model could be a future
endeavour.